Looking for a Signal in the Noise: Revisiting Obesity and the Microbiome

### 2016-06-03 submission/Sze_Obesity_mBio_2016.tex
Run make write.paper

Two recent studies have re-analyzed published data and found that when
datasets are analyzed independently there was limited support the widely
accepted hypothesis that changes in the microbiome are associated with
obesity. This hypothesis was reconsidered by increasing the number of
data sets and pooling the results across the individual datasets. The
Preferred Reporting Items for Systematic Reviews and Meta-Analyses
(PRISMA) guidelines were applied to identify 10 studies for an updated
and more synthetic analysis. Alpha diversity metrics and the relative
risk of obesity based on those metrics were used to identify a limited
number of significant associations with obesity; however, when the
results of the studies were pooled using a random effects model
significant associations were observed between Shannon diversity, number
of observed OTUs, and Shannon evenness and obesity status They were not
observed for the ratio of \emph{Bacteroidetes} and \emph{Firmicutes} or
their individual relative abundances. Although these tests yielded small
P-values, the difference between the Shannon diversity index of
non-obese and obese individuals was 2.07\%. A power analysis
demonstrated that one of the studies had sufficient power to detect a
5\% difference in diversity. When models trained on one dataset were
then tested using the other 9 datasets, the median accuracy varied
between 33.01 and 64.77\% (median=56.67\%). Although there is
statistical support for a relationship between the microbial communities
found in human feces and obesity status, this association is relatively
weak and its detection is confounded by large interpersonal variation
and insufficient sample sizes.

@@

There have been several studies that report observing a link between the
composition of microbiome and obesity in animal models and in humans.
The first such study used genetically obese mice and observed the ratio
of the relative abundances of Bacteroidetes to Firmicutes (B:F) was
lower in obese mice than lean mice (5). Translation of this result to
humans by the same researchers did not observe this effect, but did find
that obese individuals had a lower diversity than lean individuals (6).
They also showed that the relative abundance of Bacteroidetes and
Firmicutes increased and decreased, respectively, as obese individuals
lost weight while on a fat or carbohydrate restricted diet (7). Two
re-analysis studies interrogated previously published microbiome and
obesity data and concluded that the previously reported differences in
community diversity and B:F among non-obese and obese individuals could
not be generalized (8, 9). Regardless of the results using human
populations, mechanistic studies using animal models that were
manipulated with antibiotics or colonization with varied communities
appears to support the association since direct manipulation of the
communities yielded variation in animal weight (10--13). The purported
association between the differences in the microbiome and obesity have
been widely repeated with little attention given to the lack of a clear
signal in human cohort studies.

The recent publication of additional studies that collected BMI data for
each subject as well as other studies that were not included in the
earlier re-analysis studies offered the opportunity to revisit the
question relating the structure of the human microbiome to obesity
(14--22). One critique of the prior re-analysis studies is that the
authors did not aggregate the results across studies to increase the
effective sample size. It is possible that there were small associations
within each study that were not statistically significant because the
individual studies lacked sufficient power. Alternatively, diversity
metrics may mask the appropriate signal and it is necessary to measure
the association at the level of microbial populations. Walters et al.
(8) demonstrated that Random Forest machine learning models were capable
of predicting obesity status within a single cohort, but did not attempt
to test the models on other cohorts. The purpose of this study was to
perform a meta-analysis of the association between differences in the
microbiome and obesity status by analyzing and applying a more
systematic and synthetic approach than was used previously.

@@

\textbf{\emph{Literature Review and Study Inclusion.}} We followed the
Preferred Reporting Items for Systematic Reviews and Meta-Analyses
(PRISMA) guidelines to identify studies to include in our meta-analysis
(23). A detailed description of our selection process and the exact
search terms are provided in the Supplement and in figure 1. Briefly, we
searched PubMed for original research studies that involved studying
obesity and the human microbiome. The initial search yielded 187
studies. We identified number\_string{[}n{]} additional studies that
were not designed to explicitly test for an association between the
microbiome and obesity. We then manually curated the 196 studies to
select those studies that included BMI and sequence data. This yielded
10 eligible studies. An additional study was removed from our analysis
because no individuals in the study had a BMI over 30. Among the final
10 studies, 3 were from identified from our PubMed search (6, 15, 20), 5
were originally identified from the 9 studies that did not explicitly
investigate obesity but included BMI data (14, 18, 19, 24, 25), and two
datasets were used (21, 22) because these publications did not
specifically look for any metabolic or obesity conditions but had
control populations and enabled us to help mitigate against publication
biases associated with the bacterial microbiome and obesity. The
number\_string{[}n{]} studies are summarized in Tables 1 and 2.

\textbf{\emph{Sequence Analysis Pipeline.}} All sequence data were
publicly available and were downloaded from the NCBI Sequence Read
Archive, the European Nucleotide Archive, or the investigators' personal
website
(\url{https://gordonlab.wustl.edu/TurnbaughSE/_10/_09/STM/_2009.html}).
In total seven studies used 454 (6, 14, 15, 19, 20, 22, 25) and three
studies used Illumina sequencing (18, 21, 24). All of these studies used
amplification-based 16S rRNA gene sequencing. Among the studies that
sequenced the 16S rRNA gene, the researchers targeted the V1-V2 (19),
V1-V3 (14, 15, 20), V3-V5 (22, 25), V4 {[}(18); (21); {]}, and V3-4 (24)
regions. For those studies where multiple regions were sequenced, we
selected the region that corresponded to the largest number of subjects
(6, 25). We processed the 16S rRNA gene sequence data using a
standardized mothur pipeline. Briefly, our pipelines attempted to follow
previously recommended approaches for 454 and Illumina sequencing data
(26, 27). All sequences were screened for chimeras using UCHIME and
assigned to operational taxonomic units (OTUs) using the average
neighbor algorithm using a 3\% distance threshold (28, 29). All sequence
processing was performed using mothur (v.1.37.0) (30).

\textbf{\emph{Data Analysis.}} We split the overall meta-analysis into
three general strategies using R (3.3.0). First, we followed the
approach employed by Finucane et al (9) and Walters et al (8) where each
study was re-analyzed separately to identify associations between BMI
and the relative abundance of Bacteroidetes and Firmicutes, the ratio of
Bacteroidetes and Firmicutes relative abundances (B:F), Shannon
diversity, observed richness, and Shannon evenness. After each variable
was transformed to fit a normal distribution a two-tailed t-test was
performed for comparison of non-obese and obese individuals (i.e.~BMI
\textgreater{} 35.0). We performed a pooled analysis on these measured
variables using linear random effect models to correct for study effect
to asses differences on the combined dataset between non-obese and obese
groups using the lme4 (v.1.1-12) R package. Next, we compared the
community structure from non-obese and obese individuals using PERMANOVA
analysis of Bray-Curtis distance matrices. This analysis was performed
using the vegan (v.2.3-5) R package. For both analyses, the datasets
were rarefied (N=1000) so that each study within a study had the same
number of sequences. Second, for each study we partitioned the subjects
into a low or high group depending on whether their alpha diversity
metrics were below or above the median value for the study. The relative
risk (RR) was then calculated as the ratio of the number of obese
individuals in the low group to the number of obese individuals in the
high group. We then performed a Fisher exact-test to investigate whether
the RR was significantly different from 1.0 within each study and across
all of the studies using the epiR (0.9-77) and metafor (1.9-8) packages.
Third, we used the AUCRF (1.1) R package to generate Random Forest
models. For each study we developed models using either OTUs or
genus-level phylotypes. The quality of each model was assessed by
measuring the area under the curve (AUC) of the Receive Operating
Characteristic (ROC) using ten-fold cross validation. Because the
genus-level phylotype models were developed using a common reference, it
was possible to use one study's model (i.e.~the training set) to
classify the samples from the other studies (i.e.~the testing sets). The
optimum threshold for the training set was set as the probability
threshold that had the highest combined sensitivity and specificity.
This threshold was then used to calculate the accuracy of the model
applied to the test studies. To generate Receiver Operator
Characteristic (ROC) curves and calculate the accuracy of the models we
used the pROC (1.8) R package. Finally, we performed power and sample
number simulations for different effect sizes for each study using the
pwr (1.1-3) R package and base R functions. We also calculated the
actual sample size needed based on the effect size of each individual
study.

@@

\textbf{\emph{Alpha diversity analysis.}} We calculated the Shannon
diversity index, richness, and Shannon evenness, the relative abundance
of \emph{Bacteroidetes} and \emph{Firmicutes}, and the ratio of their
relative abundance (B:F) for each sample. Once we transformed each of
the six alpha diversity metrics to make them normally distributed, we
used a t-test to identify significant associations between the alpha
diversity metric and whether an individual was obese for each of the ten
studies. The B:F and the relative abundance of \emph{Firmicutes} were
not significantly associated with obesity in any study. We identified 7
P-values less than 0.05: three studies indicated obese individuals had a
lower richness, two studies indicated a significantly lower diversity,
one study indicated a significantly lower evenness, and one study
indicated a significantly higher relative abundance of
\emph{Bacteroidetes} (Figures 2 and S1). These results largely match
those of the Walters and Finucane re-analysis studies. Interestingly,
although only two of the 10 studies observed the previously reported
association between lower diversity and obesity, the other studies
appeared to have the same trend, albeit the differences were not
statistically significant. We used a random effects linear model to
combine the studies using the study as the random effect and found
statistical support for decreased richness, evenness, and diversity
among obese individuals (all P\textless{}0.011). Although there was a
significant relationship between these metrics and diversity and obesity
status, the effect size was quite small. The obese individuals averaged
7.47\% lower richness, 0.88\% lower evenness, and 2.07\% lower
diversity. There were no significant associations when we pooled the
phylum-level metrics across studies. These results indicate that obese
individuals do have a statistically significant lower diversity than
non-obese individuals; however, it is questionable whether the
difference is biologically significant.

\textbf{\emph{Relative risk.}} Building upon the alpha diversity
analysis we calculated the relative risk of being obese based on whether
an individual's alpha diversity metrics were below or above the median
metric for that study. The results using relative risk largely matched
those of using the untransformed alpha diversity data. Across the
number\_string{[}n{]} studies and six metrics, the only significant
relative risk values were the richness, evenness, and diversity values
from the Goodrich study (Figures 3 and S2). Again, although the relative
risk values were not significant for other studies, the values tended to
be above one. When we pooled the data using a random effects model, the
relative risk associated with having a richness, evenness, or diversity
below the median for the population was significantly associated with
obesity (all P\textless{}0.0044). The relative risks associated with
alpha diversity were small. The relative risk of having a low richness
was 1.30 (1.13-1.49), low evenness was 1.20 (1.06-1.37), and low
diversity was 1.27 (1.09-1.48). There were no significant difference in
the phylum-level metrics. Again, the relative risk results indicate that
individuals with a lower richness, evenness, or diversity are at
statistically significant increased risk of being obese, it is
questionable whether that risk is biologically or clinically relevant.

\textbf{\emph{Beta diversity analysis.}} Following the approach used by
the Walters and Finucane re-analysis studies, for each dataset we
calculated a Bray-Curtis distance matrix to measure the difference in
the membership and structure of the individuals from each study. We then
used PERMANOVA to test for a significant differences between the
structure of non-obese and obese individuals. The Escobar, Goodrich, and
Turnbaugh datasets indicated a significant difference in community
structure (all P\textless{}0.05). Because it was not possible to
ascertain the directionality of the difference in community structure
nor perform a pooled analysis using studies that had non-overlapping 16S
rRNA gene sequence regions it is unclear whether these differences
reflect a broader, but perhaps small, shift in community structure
between non-obese and obese individuals.

\textbf{\emph{Development of a microbiome-based classifier of obesity.}}
The Walters re-analysis study suggested that it was possible to classify
individuals as being obese or non-obese based on the composition of
their microbiota. We repeated this analysis with additional datasets
using OTU and genus-level phylotype data. For each study we developed a
Random Forest machine learning model to classify individuals. Using
ten-fold cross validation, the observed AUC values varied between 0.52
and 0.69 indicating a relatively poor ability to classify individuals
(Figure 4A). So that we could test models on other datasets, we trained
models using genus-level phylotype data for each dataset. The the
observed AUC values for the models applied to the training datasets
varied between 0.51 and 0.65, again indicating a relatively poor ability
to classify individuals from the original dataset (Figure 4B). For each
model we identified the probability where the sum of the sensitivity and
specificity was the highest. We then used this to identify a threshold
to calculate the accuracy of the models when applied to the other
number\_string{[}n-1{]} datasets (Figure 5). Although there considerable
variation in accuracy values for each model, the median accuracy for
each model varied between 0.33 (Turnbaugh) and 0.65 (HMP) (median=0.57).
When we considered the number of samples, balance of non-obese and obese
individuals, and region within the 16S rRNA gene it was not possible to
identify factors that predictably affected model performance. The
ability to predict obesity status using the relative abundance of OTUs
and genera in the communities is only marginally better than random.
These results suggest that given the large diversity of microbiome
compositions it is difficult to identify a taxonomic signal that can be
associated with obesity.

\textbf{\emph{Power and Sample Size Estimate Simulations.}} The
inability to detect a difference between non-obese and obese individuals
could be due to the lack of a true effect or because the study had
insufficient statistical power to detect a difference because of
insufficient sampling, large interpersonal variation, and unbalanced
sampling of non-obese and obese individuals. To assess this, we
calculated the power to detect differences of 1, 5, 10, and 15\% in each
of the alpha diversity metrics using the sample sizes used in each of
the studies (Figures 6, S3-S8). Although there is no biological
rationale for these effect sizes, they represent a range that is
plausible. Only the Goodrich study had power greater than 0.80 to detect
a 5\% difference in Shannon diversity and six of the studies had enough
power to detect a 10\% difference (Figure 6). None of the studies had
sufficient power to detect a 15\% difference between B:F values (Figure
S5). In fact, the maximum power among any of the studies to detect a
15\% difference in B:F values was 0.25. Among the tests for relative
risk, none of the studies had sufficient power to detect a Cohen's d of
0.10 and only two studies had sufficient power to detect a Cohen's d of
0.15. We next estimated how many individuals would need to have been
sampled to have sufficient power to detect the four effect sizes
assuming the observed interpersonal variation from each study and
balanced sampling between the two groups. To detect a 1, 5, 10, and 15\%
difference in Shannon index, the median sampling effort per group was
approximately 3,400, 140, 35, and 16 individuals, respectively. To
detect a 1, 5, 10, and 15\% difference in B:F values, the median
sampling effort per group was approximately 160,000, 6,300, 1,600, and
700 individuals, respectively. To detect a 1, 5, 10, and 15\% difference
in relative risk values using Shannon diversity, the median sampling
effort per group was approximately 39,000, 1,500, 380, and 170
individuals, respectively. These estimates indicate that most microbiome
studies are underpowered to detect modest effect sizes using either
metric. In the case of obesity, the studies were underpowered to detect
the \texttt{signif(range\_effect\_size,\ 1){[}1{]}} to 6\% difference in
diversity that was observed across the studies.

Our meta-analysis helps to provide a clarity to the ongoing debate of
whether or not there are specific microbiome-based markers that can be
associated with obesity. We performed an extensive literature review of
the existing studies on the microbiome and obesity and perform a
meta-analysis on the studies that remained based on our inclusion and
exclusion criteria. By statistically pooling the data from ten studies,
we observed significant, but small, relationships between richness,
evenness, and diversity and obesity status as well as the relative risk
of being obese based on these metrics. We also generated Random Forest
machine learning models trained on each dataset and tested on the
remaining datasets. This analysis demonstrated that the ability to
reliably classify individuals as being obese based on the composition of
their microbiome is limited. Finally, we assessed the ability of each
study to detect defined differences in alpha diversity and observed that
most studies were underpowered to detect modest effect sizes.
Considering these datasets are among the largest published, it appears
that most microbiome studies are underpowered to detect differences in
alpha diversity.

By selecting a range of possible effect sizes, we were able to
demonstrate that most studies are underpowered to detect modest
differences in alpha diversity metrics and phylum-level relative
abundances. Several factors interact to limit the power of microbiome
studies. There is wide interpersonal variation in the diversity and
structure of the human microbiome. In addition, the common experimental
designs limit their power. As we observed, most of the studies included
in our analysis were unbalanced for the variable that we were interested
in. This was also true of those studies that originally sought to
identify associations with obesity. Even with a balanced design, we
showed that it was necessary to obtain approximately 140 and 6,300
sequences per sample to detect a 5\% difference in Shannon diversity or
B:F, respectively. It was interesting that these sample sizes agreed
across studies regardless of their sequencing method, region with in the
16S rRNA gene, or subject population (Figure 6). This suggests that
regardless of the treatment or category, these sample sizes represent a
good starting point for subject recruitment when using stool samples.
Unfortunately, few studies have been published with this level of
subject recruitment. This is troubling since the positive predictive
rate of a significant finding in an underpowered study is small leading
to results that cannot be reproduced (31). Future microbiome studies
should articulate the basis for their experimental design.

>>>>

Two recent studies have re-analyzed published data and found that when
datasets are analyzed independently there was limited support the widely
accepted hypothesis that changes in the microbiome are associated with
obesity. This hypothesis was reconsidered by increasing the number of
data sets and pooling the results across the individual datasets. The
Preferred Reporting Items for Systematic Reviews and Meta-Analyses
(PRISMA) guidelines were applied to identify 10 studies for an updated
and more synthetic analysis. Alpha diversity metrics and the relative
risk of obesity based on those metrics were used to identify a limited
number of significant associations with obesity; however, when the
results of the studies were pooled using a random effects model
significant associations were observed between Shannon diversity, number
of observed OTUs, and Shannon evenness and obesity status. They were not
observed for the ratio of \emph{Bacteroidetes} and \emph{Firmicutes} or
their individual relative abundances. Although these tests yielded small
P-values, the difference between the Shannon diversity index of
non-obese and obese individuals was 2.07\%. A power analysis
demonstrated that only one of the studies had sufficient power to detect
a 5\% difference in diversity. When Random Forest machine learning
models were trained on one dataset and then tested using the other 9
datasets, the median accuracy varied between 33.01 and 64.77\%
(median=56.67\%). Although there was support for a relationship between
the microbial communities found in human feces and obesity status, this
association is relatively weak and its detection is confounded by large
interpersonal variation and insufficient sample sizes.

@@

There have been several studies that report observing a link between the
composition of microbiome and obesity in animal models and in humans.
The first such study used genetically obese mice and observed the ratio
of the relative abundances of \emph{Bacteroidetes} to \emph{Firmicutes}
(B:F) was lower in obese mice than lean mice (5). Translation of this
result to humans by the same researchers did not observe this effect,
but did find that obese individuals had a lower diversity than lean
individuals (6). They also showed that the relative abundance of
\emph{Bacteroidetes} and \emph{Firmicutes} increased and decreased,
respectively, as obese individuals lost weight while on a fat or
carbohydrate restricted diet (7). Two re-analysis studies interrogated
previously published microbiome and obesity data and concluded that the
previously reported differences in community diversity and B:F among
non-obese and obese individuals could not be generalized (8, 9).
Regardless of the results using human populations, mechanistic studies
using animal models that were manipulated with antibiotics or
colonization with varied communities appears to support the association
since these manipulation yielded variation in animal weight (10--13).
The purported association between the differences in the microbiome and
obesity have been widely repeated with little attention given to the
lack of a clear signal in human cohort studies.

The recent publication of additional studies that collected BMI data for
each subject as well as other studies that were not included in the
earlier re-analysis studies offered the opportunity to revisit the
question relating the structure of the human microbiome to obesity. One
critique of the prior re-analysis studies is that the authors did not
aggregate the results across studies to increase the effective sample
size. It is possible that there were small associations within each
study that were not statistically significant because the individual
studies lacked sufficient power. Alternatively, diversity metrics may
mask the appropriate signal and it is necessary to measure the
association at the level of microbial populations. Walters et al. (8)
demonstrated that Random Forest machine learning models were capable of
predicting obesity status within a single cohort, but did not attempt to
test the models on other cohorts. The purpose of this study was to
perform a meta-analysis of the association between differences in the
microbiome and obesity status by analyzing and applying a more
systematic and synthetic approach than was used previously.

\textbf{\emph{Literature Review and Study Inclusion.}} We followed the
Preferred Reporting Items for Systematic Reviews and Meta-Analyses
(PRISMA) guidelines to identify studies to include in our meta-analysis
(14). A detailed description of our selection process and the exact
search terms are provided in the Supplemental Text and in Figure 1.
Briefly, we searched PubMed for original research studies that involved
studying obesity and the human microbiome. The initial search yielded
187 studies. We identified number\_string{[}n{]} additional studies that
were not designed to explicitly test for an association between the
microbiome and obesity. We then manually curated the 196 studies to
select those studies that included BMI and sequence data. This yielded
10 eligible studies. An additional study was removed from our analysis
because no individuals in the study had a BMI over 30. Among the final
10 studies, 3 were from identified from our PubMed search (10, 15, 16),
5 were originally identified from the 9 studies that did not explicitly
investigate obesity but included BMI data (17--21), and two datasets
were used (22, 23) because these publications did not specifically look
for any metabolic or obesity conditions but had control populations and
enabled us to help mitigate against publication biases associated with
the bacterial microbiome and obesity. The ten studies are summarized in
Table 1.

\textbf{\emph{Alpha diversity analysis.}} We calculated the Shannon
diversity index, richness, and Shannon evenness, the relative abundance
of \emph{Bacteroidetes} and \emph{Firmicutes}, and the ratio of their
relative abundance (B:F) for each sample. Once we transformed each of
the six alpha diversity metrics to make them normally distributed, we
used a t-test to identify significant associations between the alpha
diversity metric and whether an individual was obese for each of the ten
studies. The B:F and the relative abundance of \emph{Firmicutes} were
not significantly associated with obesity in any study. We identified
seven P-values that were less than 0.05: three studies indicated obese
individuals had a lower richness, two studies indicated a significantly
lower diversity, one study indicated a significantly lower evenness, and
one study indicated a significantly higher relative abundance of
\emph{Bacteroidetes} (Figures 2 and S1). These results largely match
those of the Walters and Finucane re-analysis studies. Interestingly,
although only two of the ten studies observed the previously reported
association between lower diversity and obesity, the other studies
appeared to have the same trend, albeit the differences were not
statistically significant. We used a random effects linear model to
combine the studies using the study as the random effect and found
statistical support for decreased richness, evenness, and diversity
among obese individuals (all P\textless{}0.011). Although there was a
significant relationship between these metrics and obesity status, the
effect size was quite small. The obese individuals averaged 7.47\% lower
richness, 0.88\% lower evenness, and 2.07\% lower diversity. There were
no significant associations when we pooled the phylum-level metrics
across studies. These results indicate that obese individuals do have a
statistically significant lower diversity than non-obese individuals;
however, it is questionable whether the difference is biologically
significant.

\textbf{\emph{Relative risk.}} Building upon the alpha diversity
analysis we calculated the relative risk of being obese based on an
individual's alpha diversity metrics relative to the median metric for
that study. The results using relative risk largely matched those of
using the untransformed alpha diversity data. Across the ten studies and
six metrics, the only significant relative risk values were the
richness, evenness, and diversity values from the Goodrich study
(Figures 3 and S2). Again, although the relative risk values were not
significant for other studies, the values tended to be above one. When
we pooled the data using a random effects model, the relative risk
associated with having a richness, evenness, or diversity below the
median for the population was significantly associated with obesity (all
P\textless{}0.0044). The relative risks associated with alpha diversity
were small. The relative risk of having a low richness was 1.30
(1.13-1.49), low evenness was 1.20 (1.06-1.37), and low diversity was
1.27 (1.09-1.48). There were no significant difference in the
phylum-level metrics. Again, the relative risk results indicate that
individuals with a lower richness, evenness, or diversity are at
statistically significant increased risk of being obese, it is
questionable whether that risk is biologically or clinically relevant.

\textbf{\emph{Beta diversity analysis.}} Following the approach used by
the Walters and Finucane re-analysis studies, for each dataset we
calculated a Bray-Curtis distance matrix to measure the difference in
the membership and structure of the individuals from each study. We then
used AMOVA to test for a significant differences between the structure
of non-obese and obese individuals (Table 1). The Escobar, Goodrich, and
Turnbaugh datasets indicated a significant difference in community
structure (all P\textless{}0.05). Because it was not possible to
ascertain the directionality of the difference in community structure
nor perform a pooled analysis using studies that had non-overlapping 16S
rRNA gene sequence regions it is unclear whether these differences
reflect a broader, but perhaps small, shift in community structure
between non-obese and obese individuals.

\textbf{\emph{Development of a microbiome-based classifier of obesity.}}
The Walters re-analysis study suggested that it was possible to classify
individuals as being non-obese or obese based on the composition of
their microbiota. We repeated this analysis with additional datasets
using OTU and genus-level phylotype data. For each study we developed a
Random Forest machine learning model to classify individuals. Using
ten-fold cross validation, the observed AUC values varied between 0.52
and 0.69 indicating a relatively poor ability to classify individuals
(Figure 4A). So that we could test models on other datasets, we trained
models using genus-level phylotype data for each dataset. The
cross-validated AUC values for the models applied to the training
datasets varied between 0.51 and 0.65, again indicating a relatively
poor ability to classify individuals from the original dataset (Figure
4B). For each model we identified the probability where the sum of the
sensitivity and specificity was the highest. We then used this
probability to define a threshold for calculating the accuracy of the
models when applied to the other nine datasets (Figure 5). Although
there considerable variation in accuracy values for each model, the
median accuracy for each model varied between 0.33 (Turnbaugh) and 0.65
(HMP) (median=0.57). When we considered the number of samples, balance
of non-obese and obese individuals, and region within the 16S rRNA gene
it was not possible to identify factors that predictably affected model
performance. The ability to predict obesity status using the relative
abundance of OTUs and genera in the communities is only marginally
better than random. These results suggest that given the large diversity
of microbiome compositions it is difficult to identify a taxonomic
signal that can be associated with obesity.

\textbf{\emph{Power and Sample Size Estimate Simulations.}} The
inability to detect a difference between non-obese and obese individuals
could be due to the lack of a true effect or because the study had
insufficient statistical power to detect a difference because of
insufficient sampling, large interpersonal variation, and unbalanced
sampling of non-obese and obese individuals. To assess this, we
calculated the power to detect differences of 1, 5, 10, and 15\% in each
of the alpha diversity metrics using the sample sizes used in each of
the studies (Figures 6, S3-S8). Although there is no biological
rationale for these effect sizes, they represent a range that is
plausible. Only the Goodrich study had power greater than 0.80 to detect
a 5\% difference in Shannon diversity and six of the studies had enough
power to detect a 10\% difference (Figure 6). None of the studies had
sufficient power to detect a 15\% difference between B:F values (Figure
S5). In fact, the maximum power among any of the studies to detect a
15\% difference in B:F values was 0.25. Among the tests for relative
risk, none of the studies had sufficient power to detect a Cohen's d of
0.10 and only two studies had sufficient power to detect a Cohen's d of
0.15. We next estimated how many individuals would need to have been
sampled to have sufficient power to detect the four effect sizes
assuming the observed interpersonal variation from each study and
balanced sampling between the two groups. To detect a 1, 5, 10, or 15\%
difference in Shannon index, the median required sampling effort per
group was approximately 3,400, 140, 35, or 16 individuals, respectively.
To detect a 1, 5, 10, and 15\% difference in B:F values, the median
required sampling effort per group was approximately 160,000, 6,300,
1,600, or 700 individuals, respectively. To detect a 1, 5, 10, and 15\%
difference in relative risk values using Shannon diversity, the median
required sampling effort per group was approximately 39,000, 1,500, 380,
or 170 individuals, respectively. These estimates indicate that most
microbiome studies are underpowered to detect modest effect sizes using
either metric. In the case of obesity, the studies were underpowered to
detect the \texttt{signif(range\_effect\_size,\ 1){[}1{]}} to 6\%
difference in diversity that was observed across the studies.

Our meta-analysis helps to provide a clarity to the ongoing debate of
whether or not there are specific microbiome-based markers that can be
associated with obesity. We performed an extensive literature review of
the existing studies on the microbiome and obesity and performed a
meta-analysis on the studies that remained based on our inclusion and
exclusion criteria. By statistically pooling the data from ten studies,
we observed significant, but small, relationships between richness,
evenness, and diversity and obesity status as well as the relative risk
of being obese based on these metrics. We also generated Random Forest
machine learning models trained on each dataset and tested on the
remaining datasets. This analysis demonstrated that the ability to
reliably classify individuals as being obese based on the composition of
their microbiome was limited. Finally, we assessed the ability of each
study to detect defined differences in alpha diversity and observed that
most studies were underpowered to detect modest effect sizes.
Considering these datasets are among the largest published, it appears
that most microbiome studies are underpowered to detect differences in
alpha diversity.

By selecting a range of possible effect sizes, we were able to
demonstrate that most studies are underpowered to detect modest
differences in alpha diversity metrics and phylum-level relative
abundances. Several factors interact to limit the power of microbiome
studies. There is wide interpersonal variation in the diversity and
structure of the human microbiome. In addition, the common experimental
designs limit their power. As we observed, most of the studies included
in our analysis were unbalanced for the variable that we were interested
in. This was also true of those studies that originally sought to
identify associations with obesity. Even with a balanced design, we
showed that it was necessary to obtain approximately 140 and 6,300
sequences per sample to detect a 5\% difference in Shannon diversity or
B:F, respectively. It was interesting that these sample sizes agreed
across studies regardless of their sequencing method, region with in the
16S rRNA gene, or subject population (Figure 6). This suggests that
regardless of the treatment or category, these sample sizes represent a
good starting point for subject recruitment when using stool samples.
Unfortunately, few studies have been published with this level of
subject recruitment. This is troubling since the positive predictive
rate of a significant finding in an underpowered study is small leading
to results that cannot be reproduced (24). Future microbiome studies
should articulate the basis for their experimental design.

Two previous reviews (8, 9) have stated that there was not a consistent
association between alpha diversity and obesity; however, neither of
these studies made an attempt to pool the existing data together to try
and harness the additional power that this would give and they did not
assess whether the studies were sufficiently powered to detect a
difference. Our analysis also used 16S rRNA gene sequence data from ten
studies whereas the Finucane study used 16S rRNA gene sequence data from
three studies (7, 10, 21) and a metagenomic study (25) and the Walters
study used 16S rRNA gene sequence data from five studies (10, 15, 20,
21, 26); two studies were included in both analyses (10, 21). Our
analysis included four of these studies (10, 15, 20, 21) and excluded
three of the studies because they were too small (7), only utilized
metagenomic data (25), or used short single read Illumina HiSeq data
that has a high error rate making it untractable for \emph{de novo} OTU
clustering (26). The additional seven datasets were published after the
two reviews were performed and include datasets with more samples than
were found in the original studies. Our collection of ten studies
allowed us to largely use the same sequence analysis pipeline for all
datasets and relied heavily on the availability of public data and
access to metadata that included variables beyond the needs of the
original study. To execute this analysis, we created an automated data
analysis pipeline, which can be easily updated to add additional studies
as they become available. Similarly, it would be possible to adapt this
pipeline to other body sites and treatment or variables (e.g.~subject's
sex or age).

Similar to our study, the Walters et al (8) generated Random Forest
machine learning models to differentiate between non-obese and obese
individuals. They obtained similar AUC values to our analysis; however,
they did not attempt to test these models on the other studies in their
analysis. When we performed the inter-dataset cross validation the
median accuracy across datasets was only 56.67\% indicating that the
models did a poor job when applied to other datasets. This could be due
to differences in subject populations and methods. Considering the
median AUC for models trained and tested on the same data with ten-fold
cross validation only varied between 0.51 and 0.65 and there was not a
strong signal in the alpha diversity data, we suspect that there is
insufficient signal to reliably classify individuals.

Although we failed to find an effect this does not necessarily mean that
there is no role for the microbiome in obesity. There is strong evidence
in murine models of obesity that the microbiome and level of adiposity
can be manipulated via genetic manipulation of the animal and
manipulation of the community through antibiotics or colonizing germ
free mice with diverse fecal material from human donors (5, 10--13).
These studies appear to conflict with the observations using human
subjects. Recalling the large interpersonal variation in the structure
of the microbiome, it is possible that each individual has their own
signatures of obesity. Alternatively, it could be that the involvement
of the microbiome in obesity is at the level of a common set of
metabolites that can be produced from different structures of the
microbiome.

@@

\textbf{\emph{Sequence Analysis Pipeline.}} All sequence data were
publicly available and were downloaded from the NCBI Sequence Read
Archive, the European Nucleotide Archive, or the investigators' personal
website
(\url{https://gordonlab.wustl.edu/TurnbaughSE/_10/_09/STM/_2009.html}).
In total seven studies used 454 (6, 15, 16, 18, 20--22) and three
studies used Illumina sequencing (17, 19, 23). All of these studies used
amplification-based 16S rRNA gene sequencing. Among the studies that
sequenced the 16S rRNA gene, the researchers targeted the V1-V2 (20),
V1-V3 (15, 16, 18), V3-V5 (21, 22), V4 {[}(19); (23); {]}, and V3-4 (17)
regions. For those studies where multiple regions were sequenced, we
selected the region that corresponded to the largest number of subjects
(6, 21). We processed the 16S rRNA gene sequence data using a
standardized mothur pipeline. Briefly, our pipelines attempted to follow
previously recommended approaches for 454 and Illumina sequencing data
(27, 28). All sequences were screened for chimeras using UCHIME and
assigned to operational taxonomic units (OTUs) using the average
neighbor algorithm using a 3\% distance threshold (29, 30). All sequence
processing was performed using mothur (v.1.37.0) (31).

\textbf{\emph{Data Analysis.}} We split the overall meta-analysis into
three general strategies using R (3.3.0). First, we followed the
approach employed by Finucane et al (9) and Walters et al (8) where each
study was re-analyzed separately to identify associations between BMI
and the relative abundance of \emph{Bacteroidetes} and Firmicutes, the
ratio of \emph{Bacteroidetes} and \emph{Firmicutes} relative abundances
(B:F), Shannon diversity, observed richness, and Shannon evenness. After
each variable was transformed to fit a normal distribution a two-tailed
t-test was performed for comparison of non-obese and obese individuals
(i.e.~BMI \textgreater{} 35.0). We performed a pooled analysis on these
measured variables using linear random effect models to correct for
study effect to asses differences on the combined dataset between
non-obese and obese groups using the lme4 (v.1.1-12) R package. Next, we
compared the community structure from non-obese and obese individuals
using analysis of molecular variance (AMOVA) with Bray-Curtis distance
matrices (32). This analysis was performed using the vegan (v.2.3-5) R
package. For both analyses, the datasets were rarefied (N=1000) so that
each study within a study had the same number of sequences. Second, for
each study we partitioned the subjects into a low or high group
depending on whether their alpha diversity metrics were below or above
the median value for the study. The relative risk (RR) was then
calculated as the ratio of the number of obese individuals in the low
group to the number of obese individuals in the high group. We then
performed a Fisher exact-test to investigate whether the RR was
significantly different from 1.0 within each study and across all of the
studies using the epiR (0.9-77) and metafor (1.9-8) packages. Third, we
used the AUCRF (1.1) R package to generate Random Forest models. For
each study we developed models using either OTUs or genus-level
phylotypes. The quality of each model was assessed by measuring the area
under the curve (AUC) of the Receive Operating Characteristic (ROC)
using ten-fold cross validation. Because the genus-level phylotype
models were developed using a common reference, it was possible to use
one study's model (i.e.~the training set) to classify the samples from
the other studies (i.e.~the testing sets). The optimum threshold for the
training set was set as the probability threshold that had the highest
combined sensitivity and specificity. This threshold was then used to
calculate the accuracy of the model applied to the test studies. To
generate Receiver Operator Characteristic (ROC) curves and calculate the
accuracy of the models we used the pROC (1.8) R package. Finally, we
performed power and sample number simulations for different effect sizes
for each study using the pwr (1.1-3) R package and base R functions. We
also calculated the actual sample size needed based on the effect size
of each individual study.

@@
